---
title: "HW 4 problem set"
author: "YOUR NAME"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

For this problem set, we'll be using the packages `dplyr` and `multicon`:

```{r loading in the libraries}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  dplyr,
  multicon)
```


Lets assume that we are interested in the distribution of faculty members' heights at the University of Cincinnati. Let's create a hypothetical data frame `UCfacultyHeights` containing these values:
```{r eval=TRUE}
set.seed(1); PwomenHeight <- rnorm(2019,63.4,2.7)
set.seed(1); PmenHeight <- rnorm(2500,69.1,3.1)

PwomenCond <- rep("Women", length(PwomenHeight))
PmenCond <- rep("Men", length(PmenHeight))

Gender <- c(PwomenCond,PmenCond)
Height <- c(PwomenHeight,PmenHeight)

UCfacultyHeights <- data.frame(Gender,Height)

```

## Part 1:

#### 1. How many faculty members at UC self-identify as women? Characterize their distribution of heights (population mean, population sd). **Note that in order to get the population standard deviation you need to use the `popsd()` function from the `multicon` package.


#### 2. ... and men? 


#### Rather that having access to the entire population, you may only get access to a limited number of members of each gender. 

#### 3a. Create a sampling distribution of means for **women** and assign it to the variable name `women10` by: setting the seed at 1, then using a for-loop to take a random sample of **10 women 1,000 times**. Report the summary stats of this distribution (mean, sd).

#### 3b. Create a sampling distribution of means for **women** and assign it to the variable name `women100` by: setting the seed at 1, then using a for-loop to take a random sample of **100 women 1,000** times. Report the summary stats of this distribution (mean, sd).

#### 3c. Are there any differences in the summary stats of the two sampling distributions? Create a histogram of each distribution. Do you notice any differences here as well? Explain what may be causing these differences. (Hint in question 4).

#### 4. The SD of a sampling distribution of means is commonly referred to as _________. We typically estimate ________ from a single sample by using a central tendency measure of that sample. What is this equation. Explicitly connect this to what you observed in above.

## Part 2:

#### 5. Create a sampling distribution of means for **men** and assign it to the variable name `men50` by: setting the seed at 1, then using a for-loop to take a random sample of **50 men 100 times**. Report the summary stats of this sample (mean, sd).

#### 6. What is the largest sample mean that you get? What is the smallest?

#### 7. For the above distribution what proportion of sample means are between 69.1 and 69.3?

#### 8. Create a null distribution of means for men and assin it to the variable `nullM` by setting the seed at 1, and using the sampling aruments from question 4. What is the maximum difference you abserve assuming the null hypothesis?

#### 9. Assuming the above null distribution, what is the probablity of finding an observed difference of 2 inches between your two groups?



